---
title: "Psychometric Evaluation: Comprehensive EFA, Parallel Analysis, and Multi-Model CFA Fit Comparison"
author: "Research Team"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

knitr::opts_knit$set(root.dir = getwd())

# Define relative directory paths
data_dir   <- file.path("..", "Data")
output_dir <- file.path("..", "Data_Analysis")

generated_data_dir  <- file.path(output_dir, "generatedData")
generated_plots_dir <- file.path(output_dir, "generatedPlots")

# Create output directories if they do not exist
dir.create(generated_data_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(generated_plots_dir, recursive = TRUE, showWarnings = FALSE)

```

## Libraries

```{r libraries}
library(dplyr)
library(psych)
library(polycor)
library(lavaan)
library(knitr)
library(corrplot)
library(RColorBrewer)
library(openxlsx)

```

## Utility Functions

```{r functions}
# 1. Polychoric correlation matrix computation with column name preservation
make_corrM <- function(d = NULL, order = NULL, ...) {
  if (is.null(d)) return(NULL)
  
  d <- as.data.frame(d)
  
  # Remove items with zero variance in the current subset
  valid_cols <- sapply(d, function(col) length(unique(na.omit(col))) > 1)
  d_clean <- d[, valid_cols, drop = FALSE]
  
  if (ncol(d_clean) < 2) return(NULL)
  
  # Convert to ordered factors while strictly preserving original item names
  orig_names <- colnames(d_clean)
  d_ordered <- as.data.frame(lapply(d_clean, ordered), check.names = FALSE)
  colnames(d_ordered) <- orig_names
  
  # Compute polychoric correlation matrix using two-step approximation with pairwise deletion
  het <- tryCatch({
    polycor::hetcor(
      d_ordered, 
      use = "pairwise.complete.obs", 
      std.err = FALSE, 
      ML = FALSE
    )
  }, error = function(e) NULL)
  
  if (is.null(het)) return(NULL)
  
  corr_matrix <- het$correlations
  dimnames(corr_matrix) <- list(orig_names, orig_names)
  
  # Smooth matrix if non-positive definite eigenvalues occur in small sub-samples
  if (any(eigen(corr_matrix)$values <= 0)) {
    corr_matrix <- psych::cor.smooth(corr_matrix)
  }
  
  # Optional matrix reordering
  if (!is.null(order) && length(order) > 0) {
    valid_order <- intersect(order, colnames(corr_matrix))
    if (length(valid_order) > 1) {
      corr_matrix <- corr_matrix[valid_order, valid_order, drop = FALSE]
    }
  }
  
  return(corr_matrix)
}

# 2. Standardized extraction of CFA fit measures from lavaan objects
# 2. Standardized extraction of CFA fit measures from lavaan objects (Safe & Robust)
extract_fit_measures <- function(fit_obj, model_label) {
  # Default empty row in case of non-convergence
  empty_res <- data.frame(
    Model          = as.character(model_label),
    chisq          = NA_real_,
    df             = NA_real_,
    pvalue         = NA_real_,
    cfi            = NA_real_,
    tli            = NA_real_,
    rmsea          = NA_real_,
    rmsea.ci.lower = NA_real_,
    rmsea.ci.upper = NA_real_,
    srmr           = NA_real_,
    aic            = NA_real_,
    bic            = NA_real_,
    stringsAsFactors = FALSE
  )
  
  if (is.null(fit_obj) || !is(fit_obj, "lavaan") || !lavInspect(fit_obj, "converged")) {
    return(empty_res)
  }
  
  fm <- tryCatch({
    fitMeasures(fit_obj, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea", 
                           "rmsea.ci.lower", "rmsea.ci.upper", "srmr", "aic", "bic"))
  }, error = function(e) NULL)
  
  if (is.null(fm)) return(empty_res)
  
  # Extração segura valor por valor sem conflito de dimnames
  res_df <- data.frame(
    Model          = as.character(model_label),
    chisq          = as.numeric(fm["chisq"]),
    df             = as.numeric(fm["df"]),
    pvalue         = as.numeric(fm["pvalue"]),
    cfi            = as.numeric(fm["cfi"]),
    tli            = as.numeric(fm["tli"]),
    rmsea          = as.numeric(fm["rmsea"]),
    rmsea.ci.lower = as.numeric(fm["rmsea.ci.lower"]),
    rmsea.ci.upper = as.numeric(fm["rmsea.ci.upper"]),
    srmr           = as.numeric(fm["srmr"]),
    aic            = as.numeric(fm["aic"]),
    bic            = as.numeric(fm["bic"]),
    stringsAsFactors = FALSE
  )
  
  return(res_df)
}

```

## Factor Structure Specifications (1 to 4 Factors)

## Factor Structure Specifications (1 to 4 Factors)

```{r model definitions}
# ------------------------------------------------------------------------------
# 4-Factor Model (Granular: Visual Clarity + System Retrieval + Cognitive + Intuitiveness)
# ------------------------------------------------------------------------------
UX_factors_4f <- list(
  vis_clarity   = c('visFaithful', 'visSimpleUnderstand', 'visAppropriate'),
  sys_retrieval = c('sysRetrieveInfo', 'visRemember', 'trialError'),
  cognitive     = c('sysThinkDeep', 'sysUseful', 'sysEnjoy', 'tasksCompleteTask'),
  intuitiveness = c('sysSimpleUnderstand', 'sysEasyStart', 'analysisConfigure')
)

# ------------------------------------------------------------------------------
# 3-Factor Model (Baseline Hypothesized: Visualization + Cognitive + Intuitiveness)
# ------------------------------------------------------------------------------
UX_factors_3f <- list(
  visualization = c('visFaithful', 'visSimpleUnderstand', 'visAppropriate', 'sysRetrieveInfo', 'visRemember', 'trialError'),
  cognitive     = c('sysThinkDeep', 'sysUseful', 'sysEnjoy', 'tasksCompleteTask'),
  intuitiveness = c('sysSimpleUnderstand', 'sysEasyStart', 'analysisConfigure')
)

# ------------------------------------------------------------------------------
# 2-Factor Model A (Original: Interface Usability vs. Cognitive Utility)
# ------------------------------------------------------------------------------
UX_factors_2f_a <- list(
  interface_usability = c('visFaithful', 'visSimpleUnderstand', 'visAppropriate', 'sysRetrieveInfo', 
                          'visRemember', 'trialError', 'sysSimpleUnderstand', 'sysEasyStart', 'analysisConfigure'),
  cognitive_utility   = c('sysThinkDeep', 'sysUseful', 'sysEnjoy', 'tasksCompleteTask')
)

# ------------------------------------------------------------------------------
# 2-Factor Model B (Alternative: System Experience vs. Intuitiveness)
# ------------------------------------------------------------------------------
UX_factors_2f_b <- list(
  system_experience = c('visFaithful', 'visSimpleUnderstand', 'visAppropriate', 'sysRetrieveInfo', 
                        'visRemember', 'trialError', 'sysThinkDeep', 'sysUseful', 'sysEnjoy', 'tasksCompleteTask'),
  intuitiveness     = c('sysSimpleUnderstand', 'sysEasyStart', 'analysisConfigure')
)

# ------------------------------------------------------------------------------
# 1-Factor Model (Unidimensional Baseline)
# ------------------------------------------------------------------------------
UX_factors_1f <- list(
  general_ux = c('visFaithful', 'visSimpleUnderstand', 'visAppropriate', 'sysRetrieveInfo', 
                 'visRemember', 'trialError', 'sysThinkDeep', 'sysUseful', 'sysEnjoy', 
                 'tasksCompleteTask', 'sysSimpleUnderstand', 'sysEasyStart', 'analysisConfigure')
)

# ------------------------------------------------------------------------------
# Dynamic lavaan syntax generator
# ------------------------------------------------------------------------------
build_syntax <- function(factor_list) {
  lines <- sapply(names(factor_list), function(f_name) {
    paste0(f_name, " =~ ", paste(factor_list[[f_name]], collapse = " + "))
  })
  paste(lines, collapse = "\n")
}

syntax_1f   <- build_syntax(UX_factors_1f)
syntax_2f_a <- build_syntax(UX_factors_2f_a)
syntax_2f_b <- build_syntax(UX_factors_2f_b)
syntax_3f   <- build_syntax(UX_factors_3f)
syntax_4f   <- build_syntax(UX_factors_4f)

# ------------------------------------------------------------------------------
# Consolidate Models for CFA Testing Loop
# ------------------------------------------------------------------------------
cfa_models_list <- list(
  "1_Factor"     = syntax_1f,
  "2_Factor_A"   = syntax_2f_a,  # Usability vs Cognitive
  "2_Factor_B"   = syntax_2f_b,  # System Experience vs Intuitiveness
  "3_Factor"     = syntax_3f,
  "4_Factor"     = syntax_4f
)

# Extract all unique items for dataset filtering
columns_to_keep <- unname(unlist(UX_factors_1f))
```

## Data Loading and Preparation

```{r load data}
stimuli_letters <- c("Tableau", "JMP", "Excel", "Data Studio", "Power BI")

# Load multi-group dataset
grp_data <- read.csv(file.path(data_dir, "multigroup_ratings.csv"), encoding = "UTF-8")
data_items_only <- grp_data %>% dplyr::select(dplyr::all_of(columns_to_keep))

```

## 1. Complete Polychoric Correlation Matrices (Consolidated PDF)

```{r polychoric correlations pdf}
output_corr_pdf <- file.path(generated_plots_dir, "Polychoric_Correlations_Consolidated.pdf")

# Open PDF graphics device with dimensions suited for square matrices
pdf(output_corr_pdf, width = 11, height = 11)

# 1.1 Pooled Matrix (Full matrix display)
cat("\nGenerating Complete Polychoric Correlation Plot: POOLED_ALL\n")
c_mat_pooled <- make_corrM(data_items_only)

if (!is.null(c_mat_pooled)) {
  corrplot(
    c_mat_pooled,
    method      = "color",
    col         = COL2('RdYlBu', 100),
    type        = "full",
    order       = "original",
    addCoef.col = "black",
    number.cex  = 0.70,
    tl.col      = "black",
    tl.srt      = 45,
    tl.cex      = 0.85,
    diag        = TRUE,
    mar         = c(1, 1, 3, 1),
    title       = "Complete Polychoric Correlation Matrix - Pooled (All Systems)"
  )
}

# 1.2 Individual Matrices per Stimulus (Full matrix display)
for (stim in stimuli_letters) {
  cat("Generating Complete Polychoric Correlation Plot for stimulus:", stim, "\n")
  
  stim_data <- grp_data %>%
    dplyr::filter(stimulus == stim) %>%
    dplyr::select(dplyr::all_of(columns_to_keep)) %>%
    droplevels()
  
  if (nrow(stim_data) >= 10) {
    c_mat_stim <- make_corrM(stim_data)
    
    if (!is.null(c_mat_stim)) {
      corrplot(
        c_mat_stim,
        method      = "color",
        col         = COL2('RdYlBu', 100),
        type        = "full",
        order       = "original",
        addCoef.col = "black",
        number.cex  = 0.70,
        tl.col      = "black",
        tl.srt      = 45,
        tl.cex      = 0.85,
        diag        = TRUE,
        mar         = c(1, 1, 3, 1),
        title       = paste0("Complete Polychoric Correlation Matrix - ", stim)
      )
    }
  }
}

dev.off()

cat("\n========================================================================\n")
cat(" Consolidated Complete Polychoric Correlation Matrices saved to:\n", output_corr_pdf, "\n")
cat("========================================================================\n\n")

```

## 2. Parallel Analysis Consolidation

```{r parallel analysis consolidated}
all_parallel_results <- data.frame()

# Helper function to compute Parallel Analysis
extract_pa_row <- function(df_data, sample_name, plot_dir) {
  n_obs <- nrow(df_data)
  c_mat <- make_corrM(df_data)
  
  if (is.null(c_mat) || ncol(c_mat) < 2) return(NULL)
  
  # Export scree plot to PDF
  pdf_path <- file.path(plot_dir, paste0("ScreePlot_PA-", sample_name, ".pdf"))
  pdf(pdf_path, width = 8, height = 4.5)
  
  pa_res <- psych::fa.parallel(
    c_mat,
    n.obs  = n_obs,
    fa     = "both",
    fm     = "wls",
    n.iter = 100,
    plot   = TRUE,
    main   = paste0("Parallel Analysis Scree Plot - ", sample_name)
  )
  dev.off()
  
  row_data <- data.frame(
    Sample                  = sample_name,
    N_Observations          = n_obs,
    N_Items                 = ncol(c_mat),
    Suggested_Factors_PA    = pa_res$nfact,
    Suggested_Components_PA = pa_res$ncomp,
    
    FA_Eigen_Real_F1        = round(pa_res$fa.values[1], 3),
    FA_Eigen_Sim_F1         = round(pa_res$fa.sim[1], 3),
    FA_Eigen_Real_F2        = round(pa_res$fa.values[2], 3),
    FA_Eigen_Sim_F2         = round(pa_res$fa.sim[2], 3),
    FA_Eigen_Real_F3        = round(pa_res$fa.values[3], 3),
    FA_Eigen_Sim_F3         = round(pa_res$fa.sim[3], 3),
    FA_Eigen_Real_F4        = round(pa_res$fa.values[4], 3),
    FA_Eigen_Sim_F4         = round(pa_res$fa.sim[4], 3)
  )
  
  return(row_data)
}

# 2.1 Parallel Analysis on Pooled Data
cat("\nRunning Parallel Analysis: Pooled Data\n")
pooled_pa_row <- extract_pa_row(
  df_data     = data_items_only, 
  sample_name = "Pooled_All_Systems", 
  plot_dir    = generated_plots_dir
)
if (!is.null(pooled_pa_row)) {
  all_parallel_results <- rbind(all_parallel_results, pooled_pa_row)
}

# 2.2 Parallel Analysis by System
for (stim in stimuli_letters) {
  cat("Running Parallel Analysis for stimulus:", stim, "\n")
  
  stim_data <- grp_data %>%
    dplyr::filter(stimulus == stim) %>%
    dplyr::select(dplyr::all_of(columns_to_keep)) %>%
    droplevels()
  
  if (nrow(stim_data) >= 10) {
    stim_pa_row <- extract_pa_row(
      df_data     = stim_data, 
      sample_name = stim, 
      plot_dir    = generated_plots_dir
    )
    if (!is.null(stim_pa_row)) {
      all_parallel_results <- rbind(all_parallel_results, stim_pa_row)
    }
  } else {
    message("Insufficient sample size for stimulus: ", stim)
  }
}

knitr::kable(all_parallel_results, caption = "Consolidated Parallel Analysis Results")

```

## 3. EFA: Multi-Tab Workbook (Individual Sheets per Factor Solution and Stimulus)

```{r efa multi-tab excel}
# Master lists to hold separate tables per factor solution and per stimulus
efa_loadings_sheets_list <- list()
consolidated_variance_df <- data.frame()

# Function to run EFA models (1 to 4 factors) for a given dataset
run_efa_models <- function(df_input, label_prefix) {
  cat("\nFitting EFA Models (1-4 Factors) for:", label_prefix, "\n")
  
  c_mat <- make_corrM(df_input)
  if (is.null(c_mat)) return(list(loadings_by_factor = list(), variance = NULL))
  
  n_obs <- nrow(df_input)
  local_loadings_by_factor <- list()
  local_variance_list <- list()
  
  for (nf in 1:4) {
    rot <- if (nf == 1) "none" else "promax"
    
    efa_fit <- tryCatch({
      psych::fa(
        r        = c_mat, 
        nfactors = nf, 
        n.obs    = n_obs, 
        rotate   = rot, 
        fm       = "wls"
      )
    }, error = function(e) {
      cat("Note: EFA", nf, "factor(s) did not converge for", label_prefix, "\n")
      NULL
    })
    
    if (!is.null(efa_fit)) {
      # 1. Build factor loadings table
      raw_loadings <- unclass(efa_fit$loadings)
      item_names   <- rownames(raw_loadings)
      
      loadings_table <- data.frame(
        Item = item_names,
        stringsAsFactors = FALSE
      )
      
      for (f_idx in 1:nf) {
        f_col_name <- paste0("Loading_F", f_idx)
        loadings_table[[f_col_name]] <- round(raw_loadings[item_names, f_idx], 4)
      }
      
      loadings_table$Communality_h2 <- round(efa_fit$communality[item_names], 4)
      loadings_table$Uniqueness_u2  <- round(efa_fit$uniquenesses[item_names], 4)
      loadings_table$Complexity     <- round(efa_fit$complexity[item_names], 4)
      
      local_loadings_by_factor[[paste0(nf, "F")]] <- loadings_table
      
      # 2. Extract variance explained metrics with fixed 4 factor slots
      v_mat <- efa_fit$Vaccounted
      
      for (r_name in rownames(v_mat)) {
        var_row <- data.frame(
          Sample                   = label_prefix,
          Model_Solution           = paste0(nf, "_Factor"),
          Metric                   = r_name,
          Factor_1                 = if (ncol(v_mat) >= 1) round(v_mat[r_name, 1], 4) else NA_real_,
          Factor_2                 = if (ncol(v_mat) >= 2) round(v_mat[r_name, 2], 4) else NA_real_,
          Factor_3                 = if (ncol(v_mat) >= 3) round(v_mat[r_name, 3], 4) else NA_real_,
          Factor_4                 = if (ncol(v_mat) >= 4) round(v_mat[r_name, 4], 4) else NA_real_,
          Total_Variance_Explained = NA_real_,
          stringsAsFactors         = FALSE
        )
        
        if (r_name %in% c("Cumulative Var", "Cum.Var")) {
          var_row$Total_Variance_Explained <- round(v_mat[r_name, ncol(v_mat)], 4)
        } else if (r_name %in% c("Proportion Var", "Prop.Var", "SS loadings")) {
          var_row$Total_Variance_Explained <- round(sum(v_mat[r_name, ]), 4)
        }
        
        local_variance_list[[length(local_variance_list) + 1]] <- var_row
      }
    }
  }
  
  res_variance <- if (length(local_variance_list) > 0) dplyr::bind_rows(local_variance_list) else NULL
  
  return(list(loadings_by_factor = local_loadings_by_factor, variance = res_variance))
}

# 3.1 Execute EFA on Pooled Data
pooled_efa <- run_efa_models(data_items_only, "Pooled")
if (!is.null(pooled_efa$variance)) {
  consolidated_variance_df <- dplyr::bind_rows(consolidated_variance_df, pooled_efa$variance)
}

for (nf in 1:4) {
  nf_tag <- paste0(nf, "F")
  if (!is.null(pooled_efa$loadings_by_factor[[nf_tag]])) {
    sheet_title <- paste0(nf_tag, "_Pooled")
    efa_loadings_sheets_list[[sheet_title]] <- pooled_efa$loadings_by_factor[[nf_tag]]
  }
}

# 3.2 Execute EFA per Stimulus and Populate Dedicated Sheets
for (stim in stimuli_letters) {
  stim_data <- grp_data %>% 
    dplyr::filter(stimulus == stim) %>% 
    dplyr::select(dplyr::all_of(columns_to_keep)) %>%
    droplevels()
  
  if (nrow(stim_data) > 10) {
    stim_efa <- run_efa_models(stim_data, stim)
    if (!is.null(stim_efa$variance)) {
      consolidated_variance_df <- dplyr::bind_rows(consolidated_variance_df, stim_efa$variance)
    }
    
    clean_stim_name <- gsub(" ", "_", stim)
    
    for (nf in 1:4) {
      nf_tag <- paste0(nf, "F")
      if (!is.null(stim_efa$loadings_by_factor[[nf_tag]])) {
        sheet_title <- paste0(nf_tag, "_", clean_stim_name)
        sheet_title <- substr(sheet_title, 1, 31)
        efa_loadings_sheets_list[[sheet_title]] <- stim_efa$loadings_by_factor[[nf_tag]]
      }
    }
  }
}

# 3.3 Create and Populate Multi-Tab Excel Workbook
efa_excel_wb <- createWorkbook()

# Sheet 1: Parallel Analysis
addWorksheet(efa_excel_wb, "Parallel_Analysis")
writeData(efa_excel_wb, "Parallel_Analysis", all_parallel_results)

# Sheet 2: Consolidated Variance Explained across all solutions and samples
addWorksheet(efa_excel_wb, "Variance_Explained")
writeData(efa_excel_wb, "Variance_Explained", consolidated_variance_df)

# Sheets 3+: Dedicated Sheets for each Factor Solution x Stimulus
for (s_name in names(efa_loadings_sheets_list)) {
  addWorksheet(efa_excel_wb, s_name)
  writeData(efa_excel_wb, s_name, efa_loadings_sheets_list[[s_name]])
}

output_efa_excel <- file.path(generated_data_dir, "EFA_Summary_Consolidated.xlsx")
saveWorkbook(efa_excel_wb, output_efa_excel, overwrite = TRUE)

cat("\n========================================================================\n")
cat(" EFA Multi-Tab Workbook saved to:\n", output_efa_excel, "\n")
cat(" Total worksheets created:", length(names(efa_loadings_sheets_list)) + 2, "\n")
cat("========================================================================\n\n")

```


## 4. CFA: Multi-Tab Excel Workbook Generation (1 to 4 Factors)

```{r cfa consolidated excel}
# 4.1 Pooled CFA Models
pooled_cfa_list <- list()

for (m_name in names(cfa_models_list)) {
  m_syntax <- cfa_models_list[[m_name]]
  
  fit_pooled <- tryCatch({
    lavaan::cfa(
      model   = m_syntax,
      data    = data_items_only,
      missing = "fiml"
    )
  }, error = function(e) NULL)
  
  pooled_cfa_list[[m_name]] <- extract_fit_measures(fit_pooled, paste0("Pooled_", m_name))
}
pooled_cfa_results <- dplyr::bind_rows(pooled_cfa_list)

# 4.2 Multigroup CFA Models
multigroup_cfa_list <- list()
grp_data_clean <- grp_data %>% dplyr::filter(stimulus %in% stimuli_letters) %>% droplevels()

for (m_name in names(cfa_models_list)) {
  m_syntax <- cfa_models_list[[m_name]]
  
  fit_mg <- tryCatch({
    lavaan::cfa(
      model   = m_syntax,
      data    = grp_data_clean,
      group   = "stimulus",
      missing = "fiml"
    )
  }, error = function(e) NULL)
  
  multigroup_cfa_list[[m_name]] <- extract_fit_measures(fit_mg, paste0("Multigroup_", m_name))
}
multigroup_cfa_results <- dplyr::bind_rows(multigroup_cfa_list)

# 4.3 Individual Systems CFA Models
individual_cfa_list <- list()

for (stim in stimuli_letters) {
  stim_data <- grp_data %>% dplyr::filter(stimulus == stim) %>% droplevels()
  
  for (m_name in names(cfa_models_list)) {
    m_syntax <- cfa_models_list[[m_name]]
    
    fit_indiv <- tryCatch({
      lavaan::cfa(
        model   = m_syntax,
        data    = stim_data,
        missing = "fiml"
      )
    }, error = function(e) NULL)
    
    label_entry <- paste0(stim, "_", m_name)
    individual_cfa_list[[label_entry]] <- extract_fit_measures(fit_indiv, label_entry)
  }
}
individual_cfa_results <- dplyr::bind_rows(individual_cfa_list)

# 4.4 Build and Save Multi-Tab Excel Workbook for CFA
cfa_excel_wb <- createWorkbook()

# Tab 1: Pooled Overall
addWorksheet(cfa_excel_wb, "Pooled_Overall")
writeData(cfa_excel_wb, "Pooled_Overall", pooled_cfa_results)

# Tab 2: Multigroup
addWorksheet(cfa_excel_wb, "Multigroup")
writeData(cfa_excel_wb, "Multigroup", multigroup_cfa_results)

# Tab 3: Individual by Stimulus
addWorksheet(cfa_excel_wb, "By_Stimulus")
writeData(cfa_excel_wb, "By_Stimulus", individual_cfa_results)

output_cfa_excel <- file.path(generated_data_dir, "CFA_Model_Fit_Comparison.xlsx")
saveWorkbook(cfa_excel_wb, output_cfa_excel, overwrite = TRUE)

cat("\n========================================================================\n")
cat(" CFA Consolidated Multi-Tab Workbook saved to:\n", output_cfa_excel, "\n")
cat("========================================================================\n\n")

print(knitr::kable(pooled_cfa_results, digits = 3, caption = "CFA Fit Indices (Pooled Data Preview)"))
```

```

```